Using evolutionary algorithms to optimize the estimation of a diffusion process

### 2015-02-09 opt-methods.tex
Looking for new endeavors

\begin{abstract}
Diffusion processes are stochastic models, usually employed for modeling and forecasting in observed paths that vary along time under random environment. Such models are useful if we are able to know the values of the parameters involved. The maximum likelihood estimation of the parameters requires a system of equations that, for some cases, has no explicit solution; thus, the solution needs be estimated. We will use optimization methods to find those solutions: an Iterative Method, an algorithm based on Newton-Raphson solver, a Variable Neighbourhood Search method, a Simulated Annealing algorithm and an Evolutionary Algorithm. We generate four data sets following a Gompertz-lognormal diffusion process using different noise levels. The methods are applied to these data sets in order to estimate the parameters that rule the diffusion process. Results show that bio-inspired methods yield suitable solutions for the problem every time, even when the noise level increases. On the other hand, some analytical methods as Newton-Raphson or the Iterative Method do not always solve the problem whether their scores depend on the starting point for initial solution or the noise level hinders the resolution of the problem. In these cases, the bio-inspired algorithms remain as a suitable and
reliable approach.

@@

Many probabilistic models are frequently used for modelling natural growth-patterns and making forecast. Although there exists a lot of different growth-patterns, and many models to use, among the widely studied and useful models are diffusion processes. They are frequently employed in biological growth modelling. \cite{lognormal}.

Diffusion process is one usual mathematical formulation of a model for population growth in a randomly
varying environment \cite{lognormal}. These models also describe how a population adopts a new innovation, technology or behaviour \cite{Myrskyla2013}. With these kind of models, we have the transition density function (t.d.f.) which express the chance of observing
any size for the population at any time when the initial size is specified. With the t.d.f we also are able to determine the time at that population first reaches a fixed size. Moreover, we can forecast future values of the population size as well as confidence intervals for the predictions.

@@

A mathematical optimization or optimization problem \cite{Minoux1986} consists of maximizing or minimizing a real function by systematically choosing input values from within an allowed set and computing the value of the function.

Usually the process includes finding accurate values of some objective function given a set of constraints.

A large number of algorithms proposed for solving optimization problems can be found in bibliography, such as iterative methods that terminate in a finite number of steps, or heuristics methods that may provide approximate solutions to the problem.
However, as classical techniques of optimization might need multiple restart points and multiple runs, some authors propose using evolutionary computation \cite{Pattnaik2001} due to their population based approach.

As an example, Valdez et al. \cite{Valdez2007,Valdez2007b,Valdez2008} propose using Particle Swarm Optimization (PSO) and Genetic Algorithms (GA) to optimise a complex mathematical function, while Krishnanand et al. proposed a multimodal optimization using a PSO in \cite{Krishnanand2009}. Differential evolution (DE) has also been used as local and global selection approaches for solving multi-modal problems \cite{Ronkkonen2009}. And recently, an evolutionary multiobjective optimization (EMO) approach was proposed in \cite{Deb2010,Saha2010}.

Those optimization techniques are used to solve complex problems in the fields of, i.e., Mechanics and Engineering \cite{Tabucanon1996}, Economics, Control or Biology.
In this paper, several methods are used to estimate a Gompertz-lognormal diffusion process parameters with predictive purpose. For this case, the accuracy of the solution is very important because the output of function (\ref{funcion}) is sensitive with the input values. This is the reason because we need optimization methods that are able to generate the exact values for input parameters.

>>>>

\begin{abstract}
Diffusion processes are stochastic models, % of what? -- JJ
 usually employed for
modeling and forecasting in observed paths that vary along time under
random environment. % Don't understand -- clarify - JJ
 Such models are useful if we are able to know the values of the
 parameters involved. The maximum likelihood estimation of the
 parameters requires a system of equations that, for some cases, has
 no explicit solution; thus, the solution needs be estimated. We will
 use optimization methods to find those solutions: an Iterative
 Method, an algorithm based on Newton-Raphson solver, a Variable
 Neighbourhood Search method, a Simulated Annealing algorithm and an
 Evolutionary Algorithm. We generate four data sets following a
 Gompertz-lognormal diffusion process using different noise
 levels. % are these data sets complete? They correspond to what? - JJ
The methods are applied to these data sets in order to
 estimate the parameters that rule the diffusion process. Results show
 that bio-inspired methods yield suitable solutions for the problem in
 all four data sets, even when the noise level increases. %noise is
                                %first mentioned here. Why is it
                                %important? - JJ
 On the other hand,
 some analytical methods as Newton-Raphson or the Iterative Method do
 not always solve the problem if their scores depend on the
 starting point for initial solution or the noise level hinders the
 resolution of the problem. In these cases, the bio-inspired
 algorithms remain  a suitable and 
reliable approach.

@@

Many probabilistic models are frequently used for modelling natural
growth-patterns and making forecasts % of what? - JJ
. Although there exists a lot of
different growth-patterns, and many different ways of modeling them,
among the most widely
studied and useful models are diffusion processes. % cita - JJ
 They are frequently
employed in biological growth modelling. \cite{lognormal}.

Diffusion process is an usual mathematical formulation of a model for population growth in a randomly
varying environment \cite{lognormal}. %again same citation?  - JJ
 These models also describe how a population adopts a new innovation, technology or behaviour \cite{Myrskyla2013}. With these kind of models, we have the transition density function (t.d.f.) which express the chance of observing
any size for the population at any time when the initial size is specified. With the t.d.f we also are able to determine the time at that population first reaches a fixed size. Moreover, we can forecast future values of the population size as well as confidence intervals for the predictions.

@@

A mathematical optimization or optimization problem \cite{Minoux1986}
consists of maximizing or minimizing a real function by systematically
choosing input values from within an allowed set and computing the
value of the function.  % Er. This for an optimization journal or
                        % conference? Really? Shouldn't you start
                        % talking about diffussion processes and why
                        % they are important? - JJ

Usually the process includes finding accurate values of some objective
function given a set of constraints. % really

A large number of algorithms proposed for solving optimization
problems can be found in bibliography, % absolutely 0-information
                                % sentence - JJ
 such as iterative methods that terminate in a finite number of steps, or heuristics methods that may provide approximate solutions to the problem.
However, as classical techniques of optimization might need multiple
restart points and multiple runs, some authors propose using
evolutionary computation \cite{Pattnaik2001} due to their population
based approach. % Er. Um. This is the state of the art of what? - JJ

As an example, Valdez et al. \cite{Valdez2007,Valdez2007b,Valdez2008}
propose using Particle Swarm Optimization (PSO) and Genetic Algorithms
(GA) to optimise a complex mathematical function, while Krishnanand et
al. proposed a multimodal optimization using a PSO in
\cite{Krishnanand2009}. Differential evolution (DE) has also been used
as local and global selection approaches for solving multi-modal
problems \cite{Ronkkonen2009}. And recently, an evolutionary
multiobjective optimization (EMO) approach was proposed in
\cite{Deb2010,Saha2010}. % for what?

Those optimization techniques are used to solve complex problems in the fields of, i.e., Mechanics and Engineering \cite{Tabucanon1996}, Economics, Control or Biology.
In this paper, several methods are used to estimate a
Gompertz-lognormal diffusion process parameters with predictive
purpose. For this case, the accuracy of the solution is very important
because the output of function (\ref{funcion}) is sensitive with the
input values. This is the reason because we need optimization methods
that are able to generate the exact values for input parameters.   
% this state of the art should be rewritten - JJ
